---
title: "XGBoost fDOM Forecasting — No-Refit Workflow (Fixed Tree Ensemble)"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    theme: flatly
date: "`r Sys.Date()`"
---


## Packages and Data

Uses `tidymodels` / `parsnip` XGBoost following Melofton et al. Hyperparameters are
tuned once on pre-2024 training data via 10-fold cross-validation and random search
(`grid = 25`), and the finalized workflow is **fit once** on that same pre-2024 data.
At each forecast reference date, this fixed, already-fitted model is reused — tree
weights are never retrained. The only thing that updates at each reference date is
the initial condition: the lag input is seeded with the observed fDOM value on the
reference date, then propagated forward with the model's own predictions.

```{r packages}
library(tidyverse)
library(lubridate)
library(tidymodels)
library(vip)
tidymodels_prefer()
```


```{r load-data}
getwd()
ar_df <- read_csv("../Data/Daily_catwalk_RH_2021_2026.csv")

# Training window: all data before 2024
train_cutoff <- ymd("2024-01-01")

# Prediction window
pred_dates <- seq(ymd("2024-01-01"), ymd("2025-12-31"), by = "day")
```

```{r interpolate-short-gaps}
# Interpolate short gaps only (≤ 10 consecutive missing days) in target and all drivers.
# Gaps longer than 10 days are left as NA — those reference dates will be skipped
ar_df <- ar_df |>
  mutate(across(-Date,
                \(x) zoo::na.approx(x, maxgap = 10, na.rm = FALSE, rule = 2)))
```



## Model Fitting

### Fitting function

Fits XGBoost following Melofton et al.:

- `mtry = 0.33` (33% of features sampled per split, `counts = FALSE`)
- `trees = 1000`, `objective = "reg:squarederror"`
- `min_n`, `tree_depth`, `learn_rate`, `loss_reduction`, `sample_size` tuned via
  10-fold CV + random search (`grid = 25`)

Tuning and fitting both happen once, here, on pre-2024 data. Returns the fitted
workflow (`model`) used directly in the forecast loop — no refitting occurs there.

```{r fn-fit}
fit_xgb_model <- function(all_data,
                            train_cutoff,
                            target,
                            lag_col,
                            covariates = NULL,
                            model_name,
                            v_folds    = 10,
                            grid_size  = 25,
                            seed       = 100) {

  set.seed(seed)
  feature_cols <- c(lag_col, covariates)

  # Build training data — drop NAs in all model columns
  train_df <- all_data |>
    filter(Date < train_cutoff) |>
    select(all_of(c(target, feature_cols))) |>
    na.omit()

  message(model_name, " | n_train = ", nrow(train_df),
          " | p = ", length(feature_cols))

  # Cross-validation folds
  folds <- vfold_cv(train_df, v = v_folds)

  # XGBoost spec — mtry fixed as fraction, other hyperparameters tuned
  xgb_spec <- boost_tree(
    mtry          = 0.33,
    trees         = 1000,
    min_n         = tune(),
    tree_depth    = tune(),
    learn_rate    = tune(),
    loss_reduction = tune(),
    sample_size   = tune()
  ) |>
    set_engine("xgboost",
               objective    = "reg:squarederror",
               eval_metric  = "rmse",
               counts       = FALSE) |>
    set_mode("regression")

  # Workflow with recipe
  xgb_wf <- workflow() |>
    add_recipe(recipe(as.formula(paste(target, "~ .")), data = train_df)) |>
    add_model(xgb_spec)

  # Tune hyperparameters
  message(model_name, " | tuning hyperparameters (", v_folds, "-fold CV, grid = ",
          grid_size, ")...")

  xgb_tune <- xgb_wf |>
    tune_grid(
      resamples = folds,
      grid      = grid_size,
      control   = control_grid(save_pred = TRUE),
      metrics   = metric_set(yardstick::rmse)
    )

  best_params <- select_best(xgb_tune, metric = "rmse")
  message(model_name, " | best params selected")

  # Finalize workflow with best hyperparameters
  final_wf <- xgb_wf |> finalize_workflow(best_params)

  # Fit finalized workflow on full training data — this is the fixed model used
  # for every reference date in the forecast loop (no refitting there)
  xgb_fit <- final_wf |> fit(data = train_df)

  # In-sample fitted values (need Date for plotting — join back)
  train_df_plot <- all_data |>
    filter(Date < train_cutoff) |>
    select(Date, all_of(c(target, feature_cols))) |>
    na.omit() |>
    mutate(fitted = predict(xgb_fit, new_data = select(cur_data(), all_of(feature_cols)))$.pred)

  train_rmse <- sqrt(mean((train_df_plot[[target]] - train_df_plot$fitted)^2,
                           na.rm = TRUE))

  # Training plot
  p_fit <- ggplot(train_df_plot) +
    geom_point(aes(x = Date, y = !!sym(target), color = "Observed"),
               size = 1.2, alpha = 0.8) +
    geom_point(aes(x = Date, y = fitted, color = "XGB fitted"),
               size = 1.2, alpha = 0.8) +
    scale_color_manual(values = c("Observed" = "black", "XGB fitted" = "dodgerblue"),
                       breaks = c("Observed", "XGB fitted")) +
    labs(title = paste0(model_name, " — training fit (pre-", year(train_cutoff), ")",
                        "  |  RMSE = ", round(train_rmse, 2), " QSU"),
         x = "Date", y = "fDOM (QSU)", color = NULL) +
    theme_bw() +
    theme(legend.position = "top")

  # Variable importance
  xgb_engine <- extract_fit_engine(xgb_fit)
  imp_df <- xgboost::xgb.importance(model = xgb_engine,
                                     feature_names = feature_cols) |>
    as_tibble() |>
    rename(Variable = Feature, Importance = Gain) |>
    arrange(desc(Importance))

  p_imp <- ggplot(imp_df, aes(x = Importance, y = reorder(Variable, Importance))) +
    geom_col(fill = "#F8766D") +
    labs(title = paste0(model_name, " — variable importance"),
         x = "Importance (XGBoost gain)", y = NULL) +
    theme_bw() +
    theme(text = element_text(size = 12))

  list(
    model          = xgb_fit,     # fixed model fit once on pre-2024 data — used directly in forecast loop
    best_params    = best_params,
    plot           = p_fit,
    importance_plot = p_imp,
    model_name     = model_name,
    target         = target,
    lag_col        = lag_col,
    covariates     = covariates,
    feature_cols   = feature_cols,
    train_rmse     = train_rmse
  )
}
```

### Fit 3 models
It takes a few minutes to fit each model

```{r fit-models}
# XGB2: lag + in-lake (Diff_Dens, DO, Chla)
XGB_lake <- fit_xgb_model(
  all_data     = ar_df,
  train_cutoff = train_cutoff,
  target       = "fDOM_1_QSU_daily",
  lag_col      = "fDOM_1m_lag1",
  covariates   = c("Diff_Dens_1_max", "DOsat_1_pct_daily", "Chla_1_ugL_daily"),
  model_name   = "XGB_lag_lake"
)

# XGB3: lag + catchment (RH discharge and DOC)
XGB_catch <- fit_xgb_model(
  all_data     = ar_df,
  train_cutoff = train_cutoff,
  target       = "fDOM_1_QSU_daily",
  lag_col      = "fDOM_1m_lag1",
  covariates   = c("RH_Q_cms", "RH_DOC_mgL"),
  model_name   = "XGB_lag_catchment"
)

# XGB4: lag + in-lake + catchment
XGB_lake_catch <- fit_xgb_model(
  all_data     = ar_df,
  train_cutoff = train_cutoff,
  target       = "fDOM_1_QSU_daily",
  lag_col      = "fDOM_1m_lag1",
  covariates   = c("Diff_Dens_1_max", "DOsat_1_pct_daily", "Chla_1_ugL_daily",
                   "RH_Q_cms", "RH_DOC_mgL"),
  model_name   = "XGB_lag_lake_catchment"
)


```

### Training plots

```{r plot-training, fig.height=4}
XGB_lake$plot
XGB_catch$plot
XGB_lake_catch$plot
```

### Variable importance plots

```{r plot-importance, fig.height=4}
XGB_lake$importance_plot
XGB_catch$importance_plot
XGB_lake_catch$importance_plot
```

### Tuned hyperparameters

Best hyperparameter combination selected via cross-validated random search for each
model.

```{r print-hyperparams}
bind_rows(
  XGB_lake$best_params       |> mutate(model_name = XGB_lake$model_name, .before = 1),
  XGB_catch$best_params      |> mutate(model_name = XGB_catch$model_name, .before = 1),
  XGB_lake_catch$best_params |> mutate(model_name = XGB_lake_catch$model_name, .before = 1)
)
```

---

## 30-Day Forecast Function

The model fit in `Fit 3 models` above is used **as-is** — no refitting. The only thing
that changes at each reference date is the initial condition: the lag input is seeded
with the observed fDOM value on the reference date, then the iterated 30-day for-loop
propagates the model's own predictions forward as the lag for each next step, with
observed covariate values at each step (perfect covariate assumption).

```{r fn-forecast}
# # inputs to test function
# forecast_date <- ymd("2024-01-01")
# model_obj <- XGB_lake_catch
# all_data <- ar_df
# target  = "fDOM_1_QSU_daily"
# horizon = 30

forecast_fdom_30day <- function(forecast_date,
                                 model_obj,
                                 all_data,
                                 target  = "fDOM_1_QSU_daily",
                                 horizon = 30) {

  lag_col      <- model_obj$lag_col
  covariates   <- model_obj$covariates
  feature_cols <- model_obj$feature_cols
  model        <- model_obj$model   # fixed model fit once on pre-2024 data — not refit

  # Seed: observed fDOM on the reference date (initial condition only)
  curr_fdom <- all_data |>
    filter(Date == forecast_date) |>
    pull(!!sym(target))

  if (length(curr_fdom) == 0 || is.na(curr_fdom)) return(NULL)

  # Forecast dates: day 1 through day 30 after reference date
  forecast_dates <- seq(forecast_date + 1, forecast_date + horizon, by = "day")

  # Build full drivers dataframe upfront — observed covariates for all forecast days
  drivers <- all_data |>
    filter(Date %in% forecast_dates) |>
    select(Date, all_of(feature_cols)) |>
    arrange(Date)

  if (nrow(drivers) == 0) return(NULL)

  # Seed the lag column: first forecast step uses observed fDOM on reference date
  drivers[[lag_col]][1] <- curr_fdom

  # Step forward one day at a time using the fixed model — only the lag column updates
  pred_vals <- numeric(nrow(drivers))
  for (h in 1:nrow(drivers)) {
    temp_pred <- predict(model, new_data = drivers[h, feature_cols])$.pred
    pred_vals[h] <- temp_pred
    if (h < nrow(drivers)) {
      drivers[[lag_col]][h + 1] <- temp_pred
    }
  }

  tibble(
    model_id          = model_obj$model_name,
    reference_date    = forecast_date,
    forecast_date     = drivers$Date,
    Horizon           = as.integer(drivers$Date - reference_date),
    fDOM_1m_predicted = pred_vals
  )
}

# # test function
# zz <- forecast_fdom_30day(forecast_date = ymd("2025-01-01"), model_obj = XGB_lake_catch,
#                     all_data = ar_df, target = "fDOM_1_QSU_daily", horizon = 30)

```

---

## Run Forecasts

```{r fn-loop}
run_forecasts <- function(model_obj,
                           all_data,
                           pred_dates,
                           target  = "fDOM_1_QSU_daily",
                           horizon = 30) {

  valid_dates <- all_data |>
    filter(Date %in% pred_dates, !is.na(!!sym(target))) |>
    pull(Date)

  message(model_obj$model_name, " | forecasting ", length(valid_dates),
          " reference dates...")

  map_dfr(valid_dates, \(d) {
    cat("  -> starting forecast for ", format(d), "\n")
    forecast_fdom_30day(
      forecast_date = d,
      model_obj     = model_obj,
      all_data      = all_data,
      target        = target,
      horizon       = horizon
    )
  })
}
```

```{r run-forecasts}
forecasts_xgb2 <- run_forecasts(XGB_lake,       ar_df, pred_dates)
forecasts_xgb3 <- run_forecasts(XGB_catch,      ar_df, pred_dates)
forecasts_xgb4 <- run_forecasts(XGB_lake_catch, ar_df, pred_dates)

all_XGB_forecasts <- bind_rows(forecasts_xgb2, forecasts_xgb3, forecasts_xgb4)

glimpse(all_XGB_forecasts)
```

---

## Save Output

```{r save}
write_csv(all_XGB_forecasts,
          "../Data/XGB_forecasts_2024-2025_NoRefit_17july26.csv")


```
