---
title: "Evaluate Predictions"
author: "DWH"
date: "2026-05-22"
output: html_document
---

## Packages and Data

```{r packages}
library(tidyverse)
library(lubridate)
library(Metrics) #rmse and mae
library(patchwork)
```


```{r load-data}
#bring in observed data for eval
eval <- read_csv("../Data/Daily_catwalk_RH_2021_2026.csv") |>
  mutate(Date = as.Date(Date)) |> select(Date, fDOM_1_QSU_daily) |> 
  rename(fDOM_1m_obs = fDOM_1_QSU_daily)


na_check <- eval |> filter(is.na(fDOM_1m_obs)) |> filter(Date >= ymd("2024-01-01"))

##4 cast outputs 
AR <- read_csv("../Data/Predictions/AR_forecasts_2024-2025_RollingRefit_24july26.csv")

ARIMA <- read_csv("../Data/Predictions/ARIMA_forecasts_2024-2025_RollingRefit_24july26.csv")

XGB <- read_csv("../Data/Predictions/XGB_forecasts_2024-2025_RollingRefit_24july26.csv")

NNETAR <- read_csv("../Data/Predictions/NNETAR_forecasts_2024-2025_RollingRefit_24july26.csv")

Persistence <- read_csv("../Data/Predictions/Persistence_forecasts_2024-2025_17july26.csv")


#join predictions 
predictions <- rbind(AR, ARIMA, Persistence, XGB, NNETAR) 

unique(predictions$model_id)


#ensemble
predict_ensemble <- predictions |> 
 mutate(model_class = sub("_.*", "", model_id),
         model_vars  = sub("^[^_]*_[^_]*_", "", model_id))  |> 
  filter(model_id != "Persistence") |> 
  group_by(reference_date, forecast_date, Horizon, model_vars) |> 
  summarise(fDOM_1m_predicted = mean(fDOM_1m_predicted, na.rm = T),
            .groups = "drop") |> 
  mutate(model_id = paste0("Ensemble_lag_", model_vars)) |> 
  select(-model_vars)


predictions <- rbind(predictions, predict_ensemble) |> 
   mutate(model_vars  = sub("^[^_]*_[^_]*_", "", model_id))  |> 
    filter(model_vars != "only") |> 
    select(-model_vars)


unique(predictions$model_id)

#### join
Pred_eval_df <- left_join(predictions, eval, by = c("forecast_date" =  "Date" )) |>
  mutate(model_class = sub("_.*", "", model_id)) |>
  mutate(model_vars = sub("^[^_]*_[^_]*_", "", model_id)) |>
  mutate(model_vars = case_match(model_vars,
                                  "lake"           ~ "WaterColumn",
                                  "catchment"      ~ "Catchment",
                                  "lake_catchment" ~ "WaterColumn+Catchment",
                                  .default = model_vars)) |>
  select(model_id, model_class, model_vars,
         reference_date, forecast_date, Horizon,
         fDOM_1m_predicted, fDOM_1m_obs
         )


##set model colors
model_vars_colors <- c("WaterColumn" = "#56B4E9", "Catchment" = "#009E73", "WaterColumn+Catchment" = "#CC79A7", "Persistence" = "black")

#c("#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7", "#000000")


```



## RMSE and Skill across full TS

RMSE and Skill for ensemble: MS Figure 3

```{r}

## calc eval metrics across all models 
metrics_eval_all <- Pred_eval_df |> 
  #remove NAs and calc RMSE
  filter(!is.na(fDOM_1m_predicted), !is.na(fDOM_1m_obs)) |> 
  group_by(model_id, model_class, model_vars, Horizon) |> 
  summarise(RMSE = Metrics::rmse(fDOM_1m_obs, fDOM_1m_predicted),
            MAE = Metrics::mae(fDOM_1m_obs, fDOM_1m_predicted),
            r2 = cor(fDOM_1m_obs, fDOM_1m_predicted)^2,
            .groups = "drop"
            ) 

#pull out persistence to join for skill
rmse_persistence <- metrics_eval_all |> 
  filter(model_class == "Persistence") |> 
  select(Horizon, RMSE_persistence = RMSE)

#### RMSE across full time
rmse_fig4 <- metrics_eval_all |> 
  #filter(model_class == "AR") |> 
  filter(model_class == "Ensemble") |>   # filter(model_class %in% c("Ensemble", "Persistence")) |> 
  ggplot(aes(x = Horizon, y = RMSE, col = model_vars)) +
  geom_line(linewidth = 1.5) +
  # facet_wrap(~model_class)+
  scale_color_manual(values = model_vars_colors) +
  ylim(0.5, 3.2)+
  labs(x = "Prediction horizon (days)", y = "RMSE (QSU)", col = NULL) +
  annotate("text", x = Inf, y = Inf,  label = "Less accurate", 
           hjust = 1.1, vjust = 1.5, size = 4, fontface = "italic", color = "grey40") +
  annotate("text", x = Inf, y = -Inf, label = "More accurate", 
           hjust = 1.1, vjust = -0.5, size = 4, fontface = "italic", color = "grey40") +
  theme_bw() + theme(legend.position = "top", text = element_text(size = 18))

rmse_fig4

#### Skill across full time
skill_fig4 <- metrics_eval_all |> 
  #filter(model_class == "AR") |> 
  filter(model_class == "Ensemble") |> 
  left_join(rmse_persistence, by = "Horizon") |>
  mutate(Skill = 1 - (RMSE / RMSE_persistence)) |> 
  ggplot(aes(x = Horizon, y = Skill, color = model_vars)) +
  geom_line(linewidth = 1.5) +
  ylim(-0.2, 0.2)+
  # facet_wrap(~model_class)+
  geom_hline(yintercept = 0, linetype = "dashed", linewidth = 1.5) +
  scale_color_manual(values = model_vars_colors)+
  labs(x = "Prediction Horizon (days)", y = "Skill", color = NULL) +
  annotate("text", x = Inf, y = Inf,  label = "Better than null", 
           hjust = 1.1, vjust = 1.5, size = 4, fontface = "italic", color = "grey40") +
  annotate("text", x = Inf, y = -Inf, label = "Worse than null", 
           hjust = 1.1, vjust = -0.5, size = 4, fontface = "italic", color = "grey40") +
  theme_bw() +
  theme(legend.position = "top", text = element_text(size = 18))

skill_fig4


##join figure 4

# rmse_fig4 | skill_fig4

# ggpubr::ggarrange(rmse_fig3, skill_fig3, ncol = 2, labels = "auto")

fig4_fig <- ggpubr::ggarrange(
  rmse_fig4   + labs(x = NULL),
  skill_fig4   + labs(x = NULL),
  ncol = 2, nrow = 1,
  common.legend = TRUE, legend = "top",
  labels = c("a", "b")
)

fig4_fig <- ggpubr::annotate_figure(fig4_fig,
  bottom = ggpubr::text_grob("Prediction Horizon (days)", size = 18)
)

fig4_fig

# ggsave("../Figures/RMSE_Skill_Fig4.png", fig4_fig, width = 6, height = 4, units = "in")


##plotly figures
# plotly::ggplotly(rmse_fig4)
# plotly::ggplotly(skill_fig4)



```


RMSE and Skill by model: MS Figure 3

```{r}

#### RMSE by model
metrics_eval_all |> 
  mutate(model_class = ifelse(model_class == "NNETAR", "NNAR", model_class)) |> 
  # filter(model_class == "Ensemble") |> 
  # filter(model_class %in% c("Ensemble", "Persistence")) |> 
  ggplot(aes(x = Horizon, y = RMSE, col = model_vars)) +
  geom_line(linewidth = 1.5) +
  facet_wrap(~model_class)+
  scale_color_manual(values = model_vars_colors) +
  # ylim(0.5, 3.2)+
  labs(x = "Prediction horizon (days)", y = "RMSE (QSU)", col = NULL) +
  theme_bw() + theme(legend.position = "top", text = element_text(size = 16))

ggsave("../Figures/SI_rmse_by_structure.png", width = 7, height = 5, units = "in")


#### RMSE by model flipped presentation
# metrics_eval_all |> 
#   # filter(model_class == "Ensemble") |> 
#   # filter(model_class %in% c("Ensemble", "Persistence")) |> 
#   ggplot(aes(x = Horizon, y = RMSE, col = model_class)) +
#   geom_line(linewidth = 1.5) +
#   facet_wrap(~model_vars)+
#   # scale_color_manual(values = model_vars_colors) +
#   # ylim(0.5, 3.2)+
#   labs(x = "Prediction horizon (days)", y = "RMSE (QSU)", col = NULL) +
#   theme_bw() + theme(legend.position = "top", text = element_text(size = 18))


#### skill by model
# metrics_eval_all |> 
#   filter(model_class != "Persistence") |> 
#   left_join(rmse_persistence, by = "Horizon") |>
#   mutate(Skill = 1 - (RMSE / RMSE_persistence)) |> 
#   ggplot(aes(x = Horizon, y = Skill, color = model_vars)) +
#   geom_line(linewidth = 1.5) +
#   # ylim(-0.05, 0.25)+
#   facet_wrap(~model_class)+
#   geom_hline(yintercept = 0, linetype = "dashed", linewidth = 1.5) +
#   scale_color_manual(values = model_vars_colors)+
#   labs(x = "Prediction Horizon (days)", y = "Skill", color = NULL) +
#   theme_bw() +
#   theme(legend.position = "top", text = element_text(size = 18))



```


Identify best model across horizon by RMSE

```{r}
Pred_eval_df |> 
  #remove NAs and calc RMSE
  filter(!is.na(fDOM_1m_predicted), !is.na(fDOM_1m_obs)) |> 
  filter(model_id != "Persistence") |> 
  group_by(model_id, model_class, model_vars) |> #not grouping by horizon
  summarise(RMSE = Metrics::rmse(fDOM_1m_obs, fDOM_1m_predicted),
            .groups = "drop") |> 
  mutate(RMSE = round(RMSE, 1)) |> 
  select(-model_id) |> 
  pivot_wider(names_from = model_class, values_from = RMSE)

```



RMSE, MAE, and R2 for SI

```{r}

si_mae_r2_rmse <- metrics_eval_all |> 
  filter(model_class == "Ensemble") |> 
  pivot_longer(-c(1:4)) |> 
  # filter(model_class %in% c("Ensemble", "Persistence")) |> 
  ggplot(aes(x = Horizon, y = value, col = model_vars)) +
  geom_line(linewidth = 1.5) +
  facet_wrap(~name, nrow = 1, scales = "free_y")+
  scale_color_manual(values = model_vars_colors) +
  labs(x = "Prediction horizon (days)", y = "value", col = NULL) +
    theme_bw() + theme(legend.position = "top", text = element_text(size = 18))

# create labeller for nice facet names
metric_labels <- c(
  "RMSE" = "RMSE",
  "MAE"  = "MAE",
  "r2"   = "R^2"
)

si_mae_r2_rmse <- metrics_eval_all |> 
  filter(model_class == "Ensemble") |> 
  pivot_longer(-c(1:4)) |>
  mutate(name = factor(name, levels = c("RMSE", "MAE", "r2"))) |>  # set order
  ggplot(aes(x = Horizon, y = value, col = model_vars)) +
  geom_line(linewidth = 1.5) +
   facet_wrap(~name, nrow = 1, scales = "free_y",
             labeller = labeller(name = as_labeller(metric_labels, label_parsed)))+
  scale_color_manual(values = model_vars_colors) +
  labs(x = "Prediction horizon (days)", col = NULL) +
  theme_bw() + theme(legend.position = "top", text = element_text(size = 16))


si_mae_r2_rmse

# ggsave("../Figures/SI_rmse_mae_r2.png", si_mae_r2_rmse, width = 6, height = 4, units = "in")


# plotly::ggplotly(si_mae_r2_rmse)


```





## Absolute error over time: ensemble, 1/15/30-day horizons

```{r}

# Pred_eval_df |>
#   filter(model_class == "Ensemble", Horizon %in% c(1, 15, 30)) |>
#   filter(!is.na(fDOM_1m_predicted), !is.na(fDOM_1m_obs)) |>
#   mutate(abs_error = abs(fDOM_1m_predicted - fDOM_1m_obs),
#          Horizon   = factor(Horizon, levels = c(1, 15, 30))) |>
#   ggplot(aes(x = forecast_date, y = abs_error, color = model_vars)) +
#   geom_line(linewidth = 0.7) +
#   facet_wrap(~Horizon, ncol = 1, scales = "free_y") +
#   scale_x_date(date_breaks = "2 months", date_labels = "%b")+
#   scale_color_manual(values = model_vars_colors)+
#   labs(x = "Date", y = "Absolute Error (QSU)", color = "Horizon (days)",
#        title = "Ensemble absolute error over time, 1/15/30-day horizons") +
#   theme_bw() +
#   theme(legend.position = "top", text = element_text(size = 14))

```


## Stratfication based on Density diff

```{r}
## get strat classification
strat <- read_csv("../Data/Daily_catwalk_RH_2021_2026.csv") |> 
  select(Date, Diff_Dens_1_max) |> 
  mutate(Strat = ifelse(Diff_Dens_1_max > 0.1, "Stratified", "Mixed"))


## labels
rmse_labels <- data.frame(
  Strat = "Mixed",
  x     = c(-Inf, -Inf),
  y     = c(Inf, -Inf),
  label = c("Less accurate", "More accurate"),
  vjust = c(1.5, -0.5),
  hjust = c(-0.3, -0.3)
)

skill_labels <- data.frame(
  Strat = "Mixed",
  x     = c(Inf, Inf),
  y     = c(Inf, -Inf),
  label = c("Better than null", "Worse than null"),
  vjust = c(1.5, -0.5)
)

## calc eval metrics across all models 
metrics_eval_STRAT <- Pred_eval_df |> 
  left_join(strat, by = c("forecast_date" = "Date")) |> 
  #remove NAs and calc RMSE
  filter(!is.na(fDOM_1m_predicted), !is.na(fDOM_1m_obs),
         !is.na(Strat)) |> 
  group_by(model_id, model_class, model_vars, Horizon, Strat) |> 
  summarise(RMSE = Metrics::rmse(fDOM_1m_obs, fDOM_1m_predicted),
            MAE = Metrics::mae(fDOM_1m_obs, fDOM_1m_predicted),
            r2 = cor(fDOM_1m_obs, fDOM_1m_predicted)^2,
            .groups = "drop"
            ) 

## pull out persistence to join for skill
rmse_persistence_STRAT <- metrics_eval_STRAT |> 
  filter(model_class == "Persistence") |> 
  select(Horizon, Strat, RMSE_persistence = RMSE)

#### RMSE across full time
rmse_STRAT <- metrics_eval_STRAT |> 
  filter(model_class %in% c("Ensemble")) |> 
  ggplot(aes(x = Horizon, y = RMSE, col = model_vars)) +
  geom_line(linewidth = 1.5) +
  facet_wrap(~Strat)+
  scale_color_manual(values = model_vars_colors) +
  # ylim(0.5, 3.2)+
  labs(x = "Prediction horizon (days)", y = "RMSE (QSU)", col = NULL) +
geom_text(data = rmse_labels, aes(x = x, y = y, label = label, vjust = vjust, hjust = hjust),
          size = 4, fontface = "italic", color = "grey40", inherit.aes = FALSE)+
  theme_bw() + theme(legend.position = "top", text = element_text(size = 16))

rmse_STRAT

#### Skill across full time
skill_STRAT <- metrics_eval_STRAT |> 
  filter(model_class %in% c("Ensemble")) |> 
  left_join(rmse_persistence_STRAT, by = c("Horizon", "Strat")) |>
  mutate(Skill = 1 - (RMSE / RMSE_persistence)) |> 
  ggplot(aes(x = Horizon, y = Skill, color = model_vars)) +
  geom_line(linewidth = 1.5) +
  # ylim(-0.05, 0.25)+
  facet_wrap(~Strat)+
  geom_hline(yintercept = 0, linetype = "dashed", linewidth = 1.5) +
  scale_color_manual(values = model_vars_colors)+
  labs(x = "Prediction Horizon (days)", y = "Skill", color = NULL) +
  geom_text(data = skill_labels, aes(x = 15, y = y, label = label, vjust = vjust),
          hjust = 1.1, size = 4, fontface = "italic", color = "grey40", inherit.aes = FALSE)+
  theme_bw() +
  theme(legend.position = "top", text = element_text(size = 18))

skill_STRAT

## strats joined

rmse_STRAT | skill_STRAT

# ##plotly figures
# plotly::ggplotly(rmse_STRAT)
# plotly::ggplotly(skill_STRAT)



```



## High v low Flow 
based on USGS tinker gauge

get tinker data and join

```{r}
#change change flow class to be flow_class_usgs to use just the USGS data based classifications
flow_flags <- read_csv("../Data/HPB_USGS_Flows2.csv") |> 
   select(Date, flow_class = flow_class_hpb) 


## Join to pred eval and plot
DryWet2 <- Pred_eval_df |>
  left_join(flow_flags, by = c("forecast_date" = "Date")) |> 
  filter(!is.na(fDOM_1m_predicted), !is.na(fDOM_1m_obs),
          flow_class != "Normal") |>
  group_by(model_id, model_class, model_vars, flow_class, Horizon) |>
  summarise(RMSE = Metrics::rmse(fDOM_1m_obs, fDOM_1m_predicted),
            MAE = Metrics::mae(fDOM_1m_obs, fDOM_1m_predicted),
            r2 = cor(fDOM_1m_obs, fDOM_1m_predicted)^2,
            .groups = "drop"
            ) 

```


RMSE and skill

```{r}

#RMSE from ensemble
DryWet2_ens <- DryWet2 |> 
  #filter(model_class == "AR")
  filter(grepl("Ensemble", model_id)) 
  
flowrmse2 <- DryWet2_ens |>
  ggplot(aes(x = Horizon, y = RMSE, color = model_vars)) +
  geom_line(linewidth = 1.5) +
  facet_wrap(~flow_class) +
  labs(x = "Prediction Horizon (days)", y = "RMSE (QSU)", color = NULL) +
  scale_color_manual(values = model_vars_colors)+
  theme_bw() +  theme(legend.position = "top", text = element_text(size = 16))


flowrmse2

#### SKill
DryWet2_per <- DryWet2 |> 
  filter(grepl("Persistence", model_id)) |> 
  select(Horizon, flow_class, RMSE_persistence = RMSE)

skill_DryWet2 <- DryWet2_ens |>
  left_join(DryWet2_per, by = c("flow_class", "Horizon")) |>
  mutate(Skill = 1 - (RMSE / RMSE_persistence))

#plot
flowskill2 <- skill_DryWet2 |>
  ggplot(aes(x = Horizon, y = Skill, color = model_vars)) +
  geom_line(linewidth = 1.5) +
  geom_hline(yintercept = 0, linetype = "dashed", linewidth = 1.5) +
  facet_wrap(~flow_class) +
  # ylim(-3.5, 1)+
  labs(x = "Prediction Horizon (days)", y = "Skill", color = NULL) +
  scale_color_manual(values = model_vars_colors)+
  theme_bw() +  theme(legend.position = "top", text = element_text(size = 18))


flowskill2


#stack plots

flowrmse2 / flowskill2

# ##plotly figures
# plotly::ggplotly(flowrmse2)
# plotly::ggplotly(flowskill2)

```


## High Variability days

Set up data for high var days 

```{r}
#Firs first difference value > 90th percentile
eval2 <- eval |> mutate(fDOM_1dif = abs(fDOM_1m_obs - lag(fDOM_1m_obs, 1)))

summary(eval2$fDOM_1dif)
quantile(eval2$fDOM_1dif, probs = 0.9, na.rm = TRUE)
quantile(eval2$fDOM_1dif, probs = 0.1, na.rm = TRUE)



eval2$HighVar = ifelse(eval2$fDOM_1dif > 0.68, "High Variability", NA )
eval2$HighVar = ifelse(eval2$fDOM_1dif < 0.03, "Low Variability", eval2$HighVar )

highvar_df <- eval2 |> select(Date, HighVar)


```


```{r}
#### RMSE
rmse_HighVar <-  Pred_eval_df |> 
  left_join(highvar_df, by = c("forecast_date" = "Date")) |> 
  #filter(model_class == "AR") |> 
  filter(grepl("Ensemble", model_id))  |>
  filter(!is.na(fDOM_1m_predicted), !is.na(fDOM_1m_obs),
         !is.na(HighVar)) |>
  group_by(model_id, model_vars, model_class, HighVar, Horizon) |>
  summarise(RMSE = Metrics::rmse(fDOM_1m_obs, fDOM_1m_predicted),
            .groups = "drop"
            ) 

highvar_rmse <- rmse_HighVar |>
  ggplot(aes(x = Horizon, y = RMSE, color = model_vars)) +
  geom_line(linewidth = 1.5) +
  facet_wrap(~HighVar) +
  labs(x = "Prediction Horizon (days)", y = "RMSE (QSU)", color = NULL) +
  scale_color_manual(values = model_vars_colors)+
  theme_bw() +  theme(legend.position = "top", text = element_text(size = 16))

highvar_rmse


#### SKill
persistence_HighVar <- Pred_eval_df |>
  left_join(highvar_df, by = c("forecast_date" = "Date")) |>
  filter(model_id == "Persistence") |>
  filter(!is.na(fDOM_1m_predicted), !is.na(fDOM_1m_obs),
          !is.na(HighVar)) |>
  group_by(HighVar, Horizon) |>
  summarise(RMSE_persistence = Metrics::rmse(fDOM_1m_obs, fDOM_1m_predicted),
            .groups = "drop")

skill_HighVar <- rmse_HighVar |>
  left_join(persistence_HighVar, by = c("HighVar", "Horizon")) |>
  mutate(Skill = 1 - (RMSE / RMSE_persistence))

highvar_skill <- skill_HighVar |>
  ggplot(aes(x = Horizon, y = Skill, color = model_vars)) +
  geom_line(linewidth = 1.5) +
  geom_hline(yintercept = 0, linetype = "dashed", linewidth = 1.5) +
  facet_wrap(~HighVar) +
  ylim(-0.6,0.5)+
  labs(x = "Prediction Horizon (days)", y = "Skill", color = NULL) +
  scale_color_manual(values = model_vars_colors)+
  theme_bw() +  theme(legend.position = "top", text = element_text(size = 18))

highvar_skill


## Second version of highvar_skill with no y-axis limits set (full autoscale)
highvar_skill_full <- skill_HighVar |>
  ggplot(aes(x = Horizon, y = Skill, color = model_vars)) +
  geom_line(linewidth = 1.5) +
  geom_hline(yintercept = 0, linetype = "dashed", linewidth = 1.5) +
  facet_wrap(~HighVar) +
  labs(x = "Prediction Horizon (days)", y = "Skill", color = NULL) +
  scale_color_manual(values = model_vars_colors) +
  theme_bw() +  theme(legend.position = "top", text = element_text(size = 18))

highvar_skill_full


##plot both

highvar_rmse / highvar_skill

##plotly figures
# plotly::ggplotly(highvar_rmse)
# plotly::ggplotly(highvar_skill)

```


## Skill grouping MS figure 

```{r}

skill_STRAT / flowskill2 / highvar_skill

# (skill_STRAT / flowskill2 / highvar_skill) +
#   plot_layout(guides = "collect", axes = "collect") &
#   theme(legend.position = "top")


##arrange nicely w/ one legend, one x-axis, same panel sizes
skill_group_fig <- ggpubr::ggarrange(
  skill_STRAT   + labs(x = NULL),
  flowskill2    + labs(x = NULL),
  highvar_skill + labs(x = NULL),
  ncol = 1, nrow = 3,
  common.legend = TRUE, legend = "top",
  labels = c("a", "b", "c")
)

ggpubr::annotate_figure(skill_group_fig,
  bottom = ggpubr::text_grob("Prediction Horizon (days)", size = 18)
)

ggsave("../Figures/Skill_seasons.png", width = 7, height = 7, units = "in")

```


## RMSE grouping SI figure 

```{r}

rmse_STRAT / flowrmse2 / highvar_rmse

rmse_group_fig <- ggpubr::ggarrange(
  rmse_STRAT   + labs(x = NULL),
  flowrmse2    + labs(x = NULL),
  highvar_rmse + labs(x = NULL),
  ncol = 1, nrow = 3,
  common.legend = TRUE, legend = "top",
  labels = c("a", "b", "c")
)

ggpubr::annotate_figure(rmse_group_fig,
  bottom = ggpubr::text_grob("Prediction Horizon (days)", size = 16)
)

ggsave("../Figures/RMSE_seasons.png", width = 7, height = 7, units = "in")


```







#############################################################################################
### Forecast visualisation


### Ensemble uncertainty: `plot_forecast_uncertainty()`

Shows a single driver group's ensemble forecast as a mean line with a shaded band.
The band is the min/max range across the 4 underlying model structures (AR, ARIMA,
NNETAR, XGB) that get averaged into the ensemble for that driver group at each
forecast date -- this is calculated here rather than stored anywhere, since
`Pred_eval_df` only carries the individual model rows plus the already-averaged
`Ensemble_*` rows, not the spread across structures.

```{r fn-plot-forecast-uncertainty}
plot_forecast_uncertainty <- function(forecast_date,
                                       Pred_eval_df,
                                       obs_df,
                                       model_vars_groups = c("WaterColumn", "Catchment", "WaterColumn+Catchment"),
                                       model_classes      = c("AR", "ARIMA", "NNETAR", "XGB"),
                                       history_days       = 3,
                                       show_horizon_obs   = TRUE,
                                       show_title         = TRUE,
                                       colors             = NULL,
                                       y_limits           = NULL) {

  ref_date <- as.Date(forecast_date)

  history_obs <- obs_df |>
    filter(Date >= ref_date - history_days,
           Date <= ref_date)

  # Individual model-structure predictions feeding each driver group's ensemble
  structure_preds <- Pred_eval_df |>
    filter(reference_date == ref_date,
           model_class %in% model_classes,
           model_vars %in% model_vars_groups) |>
    select(model_class, model_vars, fc_date = forecast_date, Horizon, fDOM_1m_predicted, fDOM_1m_obs)

  if (nrow(structure_preds) == 0) {
    warning("No forecast rows found for reference_date = ", ref_date,
            " and the requested driver groups.")
    return(NULL)
  }

  # Ensemble mean + min/max envelope across model structures, per driver group
  ensemble_summary <- structure_preds |>
    group_by(model_vars, fc_date, Horizon) |>
    summarise(Ensemble_mean = mean(fDOM_1m_predicted, na.rm = TRUE),
              Ensemble_min  = min(fDOM_1m_predicted, na.rm = TRUE),
              Ensemble_max  = max(fDOM_1m_predicted, na.rm = TRUE),
              .groups = "drop")

  horizon_obs <- structure_preds |>
    distinct(fc_date, fDOM_1m_obs) |>
    filter(!is.na(fDOM_1m_obs))

  p <- ggplot() +
    geom_ribbon(data = ensemble_summary,
                aes(x = fc_date, ymin = Ensemble_min, ymax = Ensemble_max, fill = model_vars),
                alpha = 0.25, color = NA) +
    geom_point(data = history_obs,
               aes(x = Date, y = fDOM_1m_obs, shape = "Obs"),
               color = "black", size = 1.5) +
    geom_vline(xintercept = ref_date, linetype = "dashed", linewidth = 0.8) +
    geom_line(data = ensemble_summary,
              aes(x = fc_date, y = Ensemble_mean, color = model_vars),
              linewidth = 1.2) +
    labs(title = if (show_title) paste0("30-day prediction issued ", ref_date) else NULL,
         subtitle = if (show_title) "Shaded band = range across AR/ARIMA/NNETAR/XGB structures" else NULL,
         x = "Date", y = "fDOM (QSU)", color = NULL, fill = NULL, shape = NULL) +
    theme_bw() +
    theme(legend.position = "top", legend.box = "vertical", text = element_text(size = 13)) +
    guides(color = guide_legend(order = 1),
           fill  = guide_legend(order = 1),
           shape = guide_legend(order = 2))

  if (show_horizon_obs) {
    p <- p +
      geom_point(data = horizon_obs,
                 aes(x = fc_date, y = fDOM_1m_obs, shape = "Obs for Eval"),
                 color = "black", fill = "white", size = 1.5, stroke = 1.2)
  }

  p <- p +
    scale_shape_manual(values = c("Obs" = 16, "Obs for Eval" = 21),
                        breaks = c("Obs", "Obs for Eval"))

  # Same named vector (model_vars_colors) drives both the ribbon fill and the
  # mean line color, so the two legends merge into one and match the scheme
  # used everywhere else in this script
  if (!is.null(colors)) {
    p <- p + scale_color_manual(values = colors) + scale_fill_manual(values = colors)
  }

  p <- p + coord_cartesian(ylim = y_limits)

  p
}
```

```{r plot-forecast-uncertainty-example}
plot_forecast_uncertainty(
  forecast_date     = "2024-09-01",
  Pred_eval_df      = Pred_eval_df,
  obs_df            = eval,
  model_vars_groups = c("WaterColumn", "Catchment", "WaterColumn+Catchment"),
  colors            = model_vars_colors
)
```


### Model structure comparison: `plot_forecast_structures()`

Shows every model-structure line (AR/ARIMA/NNETAR/XGB) for one or more driver groups on
the same panel: color encodes the driver group (`model_vars`), line type encodes the
model structure (`model_class`).

```{r fn-plot-forecast-structures}
plot_forecast_structures <- function(forecast_date,
                                      Pred_eval_df,
                                      obs_df,
                                      model_vars_groups = c("WaterColumn", "Catchment", "WaterColumn+Catchment"),
                                      model_classes      = c("AR", "ARIMA", "NNETAR", "XGB"),
                                      history_days       = 3,
                                      show_horizon_obs   = TRUE,
                                      show_title         = TRUE,
                                      show_color_legend  = TRUE,
                                      colors             = NULL,
                                      y_limits           = NULL) {

  ref_date <- as.Date(forecast_date)

  history_obs <- obs_df |>
    filter(Date >= ref_date - history_days,
           Date <= ref_date)

  fc_preds <- Pred_eval_df |>
    filter(reference_date == ref_date,
           model_class %in% model_classes,
           model_vars %in% model_vars_groups) |>
    select(model_class, model_vars, fc_date = forecast_date, Horizon, fDOM_1m_predicted, fDOM_1m_obs) |>
    mutate(model_class = recode(model_class, "NNETAR" = "NNAR"))

  if (nrow(fc_preds) == 0) {
    warning("No forecast rows found for reference_date = ", ref_date,
            " and the requested driver groups.")
    return(NULL)
  }

  horizon_obs <- fc_preds |>
    distinct(fc_date, fDOM_1m_obs) |>
    filter(!is.na(fDOM_1m_obs))

  p <- ggplot() +
    geom_point(data = history_obs,
               aes(x = Date, y = fDOM_1m_obs, shape = "Obs"),
               color = "black", size = 1.5) +
    geom_vline(xintercept = ref_date, linetype = "dashed", linewidth = 0.8) +
    geom_line(data = fc_preds,
              aes(x = fc_date, y = fDOM_1m_predicted, color = model_vars, linetype = model_class),
              linewidth = 1) +
    labs(title = if (show_title) paste0("30-day prediction issued ", ref_date) else NULL,
         subtitle = if (show_title) "Color = driver set, line type = model structure" else NULL,
         x = "Date", y = "fDOM (QSU)", color = NULL, linetype = NULL, shape = NULL) +
    theme_bw() +
    theme(legend.position = "top", legend.box = "vertical", text = element_text(size = 13))

  # Explicit dash patterns. The built-in "dotted"/"dotdash" presets use a
  # 1-unit minimum dash segment, which some graphics devices (notably
  # Windows' default GDI-based PNG device) round away entirely, making the
  # line look solid -- custom patterns with a bigger minimum unit render
  # reliably distinct across devices
  p <- p + scale_linetype_manual(values = c("AR" = "solid", "ARIMA" = "44",
                                             "NNAR" = "23", "XGB" = "2353"))

  # At linewidth = 1, the dash/dot pattern's gaps are so small relative to the
  # dot size that dotted/dotdash read as solid in the legend key -- override
  # just the key glyph's linewidth (thinner) so the pattern stays legible;
  # the actual plotted lines are unaffected
  if (show_color_legend) {
    p <- p + guides(color = guide_legend(order = 1),
                     linetype = guide_legend(order = 2, keywidth = unit(1.4, "cm"),
                                              override.aes = list(linewidth = 0.6)),
                     shape = guide_legend(order = 3))
  } else {
    p <- p + guides(color = "none",
                     linetype = guide_legend(order = 1, keywidth = unit(1.4, "cm"),
                                              override.aes = list(linewidth = 0.6)),
                     shape = guide_legend(order = 2))
  }

  if (show_horizon_obs) {
    p <- p +
      geom_point(data = horizon_obs,
                 aes(x = fc_date, y = fDOM_1m_obs, shape = "Obs for Eval"),
                 color = "black", fill = "white", size = 1.5, stroke = 1.2)
  }

  p <- p +
    scale_shape_manual(values = c("Obs" = 16, "Obs for Eval" = 21),
                        breaks = c("Obs", "Obs for Eval"))

  if (!is.null(colors)) {
    p <- p + scale_color_manual(values = colors)
  }

  p <- p + coord_cartesian(ylim = y_limits)

  p
}
```

```{r plot-forecast-structures-example}
plot_forecast_structures(
  forecast_date = "2024-09-01",
  Pred_eval_df  = Pred_eval_df,
  obs_df        = eval,
  colors        = model_vars_colors
)

# Single driver group: linetype alone distinguishes the 4 model structures
plot_forecast_structures(
  forecast_date     = "2024-09-01",
  Pred_eval_df      = Pred_eval_df,
  obs_df            = eval,
  model_vars_groups = "WaterColumn",
  colors            = model_vars_colors
)
```

### Uncertainty + structures, side by side

No titles, shared/collected legend so both panels come out the same size and the
legend rows line up at the same height.

```{r plot-forecast-uncertainty-structures-combined}
p_uncertainty <- plot_forecast_uncertainty(
  forecast_date     = "2024-09-01",
  Pred_eval_df      = Pred_eval_df,
  obs_df            = eval,
  model_vars_groups = c("WaterColumn", "Catchment", "WaterColumn+Catchment"),
  colors            = model_vars_colors,
  show_title        = FALSE
)

p_structures <- plot_forecast_structures(
  forecast_date      = "2024-09-01",
  Pred_eval_df       = Pred_eval_df,
  obs_df             = eval,
  colors             = model_vars_colors,
  model_vars_groups  = "Catchment",
  show_title         = FALSE,
  show_color_legend  = FALSE
)

# match y-axis range across both panels
y_range <- range(c(layer_scales(p_uncertainty)$y$range$range,
                    layer_scales(p_structures)$y$range$range))

p_uncertainty <- p_uncertainty + coord_cartesian(ylim = y_range)
p_structures  <- p_structures  + coord_cartesian(ylim = y_range)

(p_structures | p_uncertainty) +
  plot_layout(guides = "collect") +
  plot_annotation(tag_levels = "a", tag_suffix = ")") &
  theme(legend.position = "top", legend.box = "vertical",
        legend.margin = margin(0, 0, 0, 0),
        legend.box.margin = margin(0, 0, 0, 0),
        legend.box.spacing = unit(2, "pt"),
        legend.spacing.y = unit(0, "pt"),
        legend.key.size = unit(0.4, "cm"),
        legend.text = element_text(size = 10),
        legend.title = element_text(size = 10),
        plot.margin = margin(2, 2, 2, 2),
        plot.tag = element_text(size = 12, margin = margin(0, 4, 0, 0)))
```






